---
title: ""
format: pdf
---

Group #28 

**Group: Mortality Explorers** 

**Group Members: Ndey Isatou Jobe (SM Biostatistics), Khondoker Nazmoon Nabi ( PhD Biostatistics), Bennet Cho (MPH Clinical Effectiveness Program)**

HW #4: Project Check-In

10/8/23

We have read and attest to the statement in the BST 210 HW assignment.

## Q1.

\"Under 5 mortality\". Using the World Bank database, we will look at associations between CO2 emissions and mortality in children under 5 years of age controlling for socioeconomic factors. 

## Q2.

Data is sourced from the World Bank DataBank. Data is collected on all countries of the world, on yearly CO2 emissions, yearly under 5 mortality rate, and other variables to account for confounding and effect measurement.

## Q3.

-   What is the relationship between a country\'s CO2 emissions and under-5 mortality rates?

-   Is there an association between higher income and lower under-5 mortality rates?

## Q4.

Does the relationship between CO2 emissions and under-5 mortality vary across different countries?

## Q5.

Under-5 mortality (Continuous)

## Q6.

-   Data Preprocessing:

Clean and preprocess the data to investigate missing values, outliers, and inconsistencies.

-   Data Exploration:

Conduct exploratory data analysis (EDA) to understand the distribution of our variables and identify potential patterns or outliers.

-    Hypothesis Formulation:

Formulate our hypotheses that higher CO2 emissions are associated with higher under-5 mortality rates, and that higher income is associated with lower mortality rates.

-   Correlation Analysis:

Begin plotting scatter plots to assess individual relationships and by calculating correlation coefficients between CO2 emissions, income, and under-5 mortality rates. Use statistical tests to assess the significance of these correlations.

-   Regression Analysis:

Perform regression analysis to examine the relationships more thoroughly. Use multiple regression to account for the influence of income while assessing the relationship between CO2 emissions and mortality rates. 

-   Interpret the coefficients (β1 and β2) and assess their statistical significance:

Interpret the results of our regression analysis, focusing on the relationship between CO2 emissions, income, and under-5 mortality rates. Discuss whether the relationships are statistically significant and the magnitude of their effects.

-   Stratified Analysis:

investigate the relationship across regions, consider stratified analysis. Analyse the relationship between CO2 emissions, income, and mortality rates separately for different regions or groups of countries.

-   Interaction Effects:

Assess whether there are interaction effects between CO2 emissions and income that influence mortality rates differently in various regions. Include interaction terms in our regression model if appropriate.

-   Model Evaluation:

Evaluate the goodness of fit of your regression models, such as adjusted R-squared, AIC, MSE, and check for model assumptions, including linearity, normality of residuals, and homoscedasticity.

-    Interpretation of Findings:

Interpret the results of your regression analysis, focusing on the relationship between CO2 emissions, income, and under-5 mortality rates. Discuss whether the relationships are statistically significant and the magnitude of their effects.

-   Regional Comparisons:

Compare the findings across regions to determine if the relationships hold consistently or if there are regional variations.Q7.

## Q7.

Our biggest challenge foreseen is how to collectively agree on the best model, as modelling is art and subjective

## Q8.

Boston Children\'s Hospital Global Health research team: <https://www.childrenshospital.org/programs/global-health/meet-our-team>

## Q9.

We will be using the R software package to complete this task.

## Q10.

```{r, eval=TRUE, echo=FALSE}
#Load packages
library(readxl)
library(dplyr)
```

# Data Exploration

```{r, eval=TRUE, echo=FALSE}
# Load data
data0 <- read_xlsx("P_Data_Extract_From_Sustainable_Development_Goals_(SDGs) (3) (1).xlsx")
attach(data0)
View(data0)
```

```{r, eval=TRUE, echo=FALSE}
# Data Wrangling
data0 <- data0 |> rename(female_employment = `Contributing family workers, female (% of female employment) (modeled ILO estimate) [SL.FAM.WORK.FE.ZS]`)

data0 <- data0 |> rename(male_employment = `Contributing family workers, male (% of male employment) (modeled ILO estimate) [SL.FAM.WORK.MA.ZS]`)


```

```{r, eval=TRUE, echo=FALSE}
#Data exploration
# View the structure of the dataset
str(data0)

# Summary statistics
summary(data0)
```

## Data Visualization

```{r, eval=TRUE, echo=FALSE}
hist(`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`)
```

```{r, eval=TRUE, echo=FALSE}
hist(log(`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`))
```

```{r, eval=TRUE, echo=FALSE}
hist(`CO2 emissions (kg per PPP $ of GDP) [EN.ATM.CO2E.PP.GD]`)
```

```{r, eval=TRUE, echo=FALSE}
hist(`CO2 emissions (metric tons per capita) [EN.ATM.CO2E.PC]`)
```

```{r, eval=TRUE, echo=FALSE}
hist(`GDP per capita (current US$) [NY.GDP.PCAP.CD]`)
```

```{r, eval=TRUE, echo=FALSE}
# Data Visualization
scatter.smooth(data0$`CO2 emissions (kg per PPP $ of GDP) [EN.ATM.CO2E.PP.GD]`, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$`CO2 emissions (metric tons per capita) [EN.ATM.CO2E.PC]`, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$`GDP per capita (current US$) [NY.GDP.PCAP.CD]`, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$`Compulsory education, duration (years) [SE.COM.DURS]`, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$`Prevalence of current tobacco use, males (% of male adults) [SH.PRV.SMOK.MA]`, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$`Prevalence of current tobacco use, females (% of female adults) [SH.PRV.SMOK.FE]`, data0$`Mortality rate, under-5, male (per 1,000 live births) [SH.DYN.MORT.MA]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$`People practicing open defecation (% of population) [SH.STA.ODFC.ZS]`, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$`Prevalence of moderate or severe food insecurity in the population (%) [SN.ITK.MSFI.ZS]`, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$female_employment, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$male_employment, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

```{r, eval=TRUE, echo=FALSE}
scatter.smooth(data0$`People using at least basic drinking water services (% of population) [SH.H2O.BASW.ZS]`, data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, ylab = "Under 5 Mortality Rate per 1,000 Live Births")
```

## Q11.

Project Attestation:  No member of this group is using these data or same/similar questions in any other course or course project, at HSPH.

## APPENDIX

```{r, eval=FALSE, echo=TRUE}
#Load packages
library(readxl)
#Load packages
library(readxl)

# Load data
data0 <- read_xlsx("P_Data_Extract_From_Sustainable_Development_Goals_(SDGs) 
                   (3) (1).xlsx")
attach(data0)

#Data exploration
# View the structure of the dataset
str(data0)

# Summary statistics
summary(data0)

# Data Visualizations
hist(`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`)

hist(log(`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`))

hist(`CO2 emissions (kg per PPP $ of GDP) [EN.ATM.CO2E.PP.GD]`)

hist(`CO2 emissions (metric tons per capita) [EN.ATM.CO2E.PC]`)

hist(`GDP per capita (current US$) [NY.GDP.PCAP.CD]`)

scatter.smooth(data0$`CO2 emissions (kg per PPP $ of GDP) 
               [EN.ATM.CO2E.PP.GD]`, data0$`Mortality rate, 
               under-5 (per 1,000 live births) [SH.DYN.MORT]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")


scatter.smooth(data0$`CO2 emissions (metric tons per capita) [EN.ATM.CO2E.PC]`, 
               data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")

scatter.smooth(data0$`GDP per capita (current US$) [NY.GDP.PCAP.CD]`, 
               data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")


scatter.smooth(data0$`Compulsory education, duration (years) [SE.COM.DURS]`, 
               data0$`Mortality rate, 
               under-5 (per 1,000 live births) [SH.DYN.MORT]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")

scatter.smooth(data0$`Prevalence of current tobacco use, 
               males (% of male adults) [SH.PRV.SMOK.MA]`, 
               data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")

scatter.smooth(data0$`Prevalence of current tobacco use,
               females (% of female adults) [SH.PRV.SMOK.FE]`, 
               data0$`Mortality rate, under-5, male (per 1,000 live births) 
               [SH.DYN.MORT.MA]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")

scatter.smooth(data0$`People practicing open defecation 
               (% of population) [SH.STA.ODFC.ZS]`, 
               data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")

scatter.smooth(data0$`Prevalence of moderate or severe 
               food insecurity in the population (%) 
               [SN.ITK.MSFI.ZS]`, data0$`Mortality rate, 
               under-5 (per 1,000 live births) [SH.DYN.MORT]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")

scatter.smooth(data0$`Contributing family workers, female
               (% of female employment) (modeled ILO estimate) [SL.FAM.WORK.FE.ZS]`, 
               data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")

scatter.smooth(data0$`Contributing family workers, 
               male (% of male employment) (modeled ILO estimate) [SL.FAM.WORK.MA.ZS]`, 
               data0$`Mortality rate, under-5 (per 1,000 live births) [SH.DYN.MORT]`, 
               
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")

scatter.smooth(data0$`People using at least basic drinking water services 
               (% of population) [SH.H2O.BASW.ZS]`, 
               data0$`Mortality rate, under-5 (per 1,000 live births) 
               [SH.DYN.MORT]`, 
               ylab = "Under 5 Mortality Rate per 1,000 Live Births")


```
